---
title: "Amy Y"
format: html
---

```{r}
library(ggplot2)
library(tidyverse)
library(sf)
library(rnaturalearth)

estimated_numbers <- read.csv("../../data/archive 2/estimated_numbers.csv") 
incidence <-read.csv("../../data/archive 2/incidence_per_1000_pop_at_risk.csv")
reported <-read.csv("../../data/archive 2/reported_numbers.csv") #use this one

world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
```

```{r}
dim(reported)

reported_clean <- reported %>%
  rename(deaths = No..of.deaths)   

map_data <- world %>%
  left_join(reported_clean, by = c("name" = "Country"))
```

```{r}
ggplot(map_data) +
  geom_sf(aes(fill = deaths), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", trans = "log10") +
  facet_wrap(~ Year) +
  theme_minimal() +
  labs(
    title = "Reported Deaths by Country Over Time",
    fill = "Deaths (log scale)"
  )
```

##RQ: reported deaths vs time vs country

```{r}
library(plotly)
library(dplyr)
library(countrycode)
```

```{r}
# Example: cleaning reported data
# Replace 'country', 'year', and 'deaths' with your real column names
reported_clean_2 <- reported %>%
  mutate(
    country = countrycode(Country, origin = "country.name", destination = "iso3c"),
  #  year = as.integer(Year),
 #   deaths = as.numeric(No..of.deaths)
  ) %>%
  filter(!is.na(country), 
        # !is.na(year), 
       #  !is.na(deaths)
         )
```

```{r}
fig <- plot_ly(
  data = reported_clean_2,
  type = "choropleth",
  locations = ~country,
  z = ~No..of.deaths,
  frame = ~Year,
  colorscale = "Reds",
  text = ~paste(Country, "<br>Deaths:", No..of.deaths),
  colorbar = list(title = "Deaths"),
  reversescale = FALSE
)

fig <- fig %>%
  layout(
    title = "Reported Deaths by Country Over Time",
    geo = list(showframe = FALSE,
               showcoastlines = FALSE,
               showcountries = TRUE,
               countrycolor = "gray",
               projection = list(type = "equirectangular"),
               bgcolor = "white",
               landcolor = "white")
  )

fig
```

::: {.callout-note title="Caption"}
This interactice choropleth map shows the number of reported deaths of malaria by countries from 2000 to 2017. Countries shaded in deeper red tones indicate higher death counts, while countries in paler or base colors report fewer or no deaths in a given year. All the coutries appearing with the base color are included in the dataset. 
:::

##RQ: total case vs deaths vs time vs country

```{r}
malaria_clean <- reported %>%
  select(Country, Year, `No..of.cases`) %>%
  rename(
    country = Country,
    year = Year,
    cases = `No..of.cases`
  )

```

```{r}
library(plotly)
fig <- plot_ly(
  data = malaria_clean,
  locations = ~country,
  locationmode = 'country names',
  z = ~cases,
  text = ~paste(country, "<br>Cases:", cases),
  type = 'choropleth',
  frame = ~year,
  colorscale = 'Reds',
  marker = list(line = list(color = "gray", width = 0.5))
)

fig <- fig %>%
  layout(
    title = "Estimated Malaria Cases Over Time (Median)",
    geo = list(
      showframe = FALSE,
      showcoastlines = FALSE
    ),
    coloraxis = list(colorbar = list(title = "Cases"))
  )

fig <- fig %>%
  layout(
    title = "Estimated Malaria Cases and Deaths Over Time (Median)",
    geo = list(
      showframe = FALSE,
      showcoastlines = FALSE,
      showcountries = TRUE,
               countrycolor = "gray",
               projection = list(type = "equirectangular"),
               bgcolor = "white",
               landcolor = "white"
    ),
    coloraxis = list(
      colorbar = list(title = "Cases")
    )
  )


fig

```
# Explanation
This map visualizes the number of reported malaria-related deaths by country from the year 2000 to 2017. Each country is colored on a gradient from light to dark red, where darker shades indicate a higher number of deaths. As the animation progresses through time, we can observe how the burden of malaria shifts across regions. Notably, countries in sub-Saharan Africa—especially the Democratic Republic of Congo and Nigeria—consistently show the darkest shades, highlighting their persistent struggle with malaria mortality. This temporal and spatial visualization helps emphasize where intervention efforts have been or still need to be concentrated.



```{r}
malaria_clean <- reported %>%
  select(Country, Year, `No..of.cases`, `No..of.deaths`) %>%
  rename(
    country = Country,
    year = Year,
    cases = `No..of.cases`,
    deaths = `No..of.deaths`
  )


fig <- plot_ly() %>%
  add_trace(
    data = malaria_clean,
    type = "choropleth",
    locations = ~country,
    locationmode = "country names",
    z = ~cases,
    frame = ~year,
    colorscale = "Blues",
    reversescale = TRUE,
    text = ~paste("Country:", country,
                  "<br>Total Cases:", format(cases, big.mark = ","),
                  "<br>Deaths:", format(deaths, big.mark = ",")),
    hoverinfo = "text",
    colorbar = list(title = "Total Cases"),
    marker = list(line = list(color = "white", width = 0.5))
  ) %>%
  

  add_trace(
    data = malaria_clean,
    type = "scattergeo",
    locations = ~country,
    locationmode = "country names",
    frame = ~year,
    marker = list(
      size = ~sqrt(deaths) * 0.25,
      color = 'red',
      opacity = 0.6,
      line = list(width = 0)
    ),
    hoverinfo = "none",
    showlegend = FALSE
  ) %>%
  
  layout(
    title = "Global Malaria Cases and Deaths from 2000 to 2017",
    geo = list(
      showframe = FALSE,
      showcountries = TRUE,
      showcoastlines = FALSE,
      countrycolor = "grey",
      landcolor = "lightgrey",
      projection = list(type = "equirectangular"),
      bgcolor = "white"
    )
  )

fig

```
# Explanation

This interactive map illustrates global malaria cases and deaths from the year 2000 to 2017. The background color of each country represents the total number of malaria cases, using a blue gradient—darker blue indicates higher case counts. Overlaying the map, we use red circles to represent the number of reported deaths. The larger the circle, the higher the death toll.

This dual-encoded map lets us compare both the prevalence of malaria and its mortality burden at once. For instance, you can see that many central and western African countries not only have dark blue shading—indicating high case numbers—but also large red circles, revealing significant death counts. Meanwhile, other regions may have lighter shades with smaller circles, showing lower disease burden.

Overall, this visualization helps us track the geographic and temporal patterns of malaria, highlighting where the disease hits hardest and where public health efforts should be prioritized.
