---
title: "Malaria Cases: 2000-2017"
execute: 
  echo: true      # change to true to show the code
code-fold: false  # change to true to fold the code chunks
---

<<<<<<< Updated upstream
# Introduction

Malaria is a life-threatening disease caused by infected female *Anopheles* mosquitoes. This analysis explores malaria cases from 2000 to 2017 globally and by regions (Americas, Africa, Eastern Mediterranean, Europe, South-East Asia, Western Pacific), focusing on trends in case numbers and death rates.

# Research Question, Aims, and Methodology

## Research Question

**How has global malaria cases and reported death cases changed over time and across different regions from 2000 to 2017?**

## Aims

-   To visualize trends in malaria cases and deaths globally from 2000 onwards.
-   To analyze regional differences and identify patterns or anomalies over time.
-   To better understand where malaria remains a major public health challenge.

## Methodology

-   **Dataset:** WHO malaria data (reported cases and deaths by country and year).
-   **Visualization Methods:**
    -   Scatter plots to show temporal trends.
    -   Dynamic choropleth maps to demonstrate spatial changes.

# Dataset Overview

-   **Total Rows:** 1,944
-   **Total Columns:** 5
    -   `Country` (Categorical): 190 unique country names.
    -   `Year` (Numerical): Reporting year (2000–2017).
    -   `No. of cases` (Numerical): Reported malaria cases (0 to 15.1 million).
    -   `No. of deaths` (Numerical): Reported malaria deaths (0 to 39,054).
    -   `WHO Region` (Categorical): 6 regions (Africa, Americas, Eastern Mediterranean, Europe, South-East Asia, Western Pacific).

# Results

## Total Reported Malaria Cases Over Time

```{r malaria-cases-over-time}
# Insert your code for plotting total malaria cases over time
```

## Distribution of Reported Malaria Deaths by Year (Global)

```{r malaria-deaths-global}
# Insert your code for plotting global malaria deaths distribution by year
```

## Reported Malaria Deaths Over Time by Region

```{r malaria-deaths-region}
# Insert your code for plotting deaths over time by region
```

## Total Reported Malaria Deaths Over Time by Region

```{r total-deaths-region}
# Insert your code for plotting total deaths by region over time
```

## Reported Cases by Country Over Time

```{r cases-by-country}
# Insert your code for plotting cases by country
```

## Reported Deaths by Country Over Time

```{r deaths-by-country}
# Insert your code for plotting deaths by country
```

## Malaria Cases and Deaths by Country Over Time

```{r cases-deaths-country}
# Insert your code for a combined plot of cases and deaths over time by country
```

# Future Work and Questions

-   Explore causes behind regional differences (e.g., healthcare access, climate, interventions).
-   Update the dataset with more recent years.
-   Investigate specific country case studies for deeper insights.
-   Analyze the relationship between malaria interventions and case/death trends.

# Conclusion

Teamwork and visual analytics have enabled a clearer understanding of malaria's global impact. Future studies should continue to leverage updated data and innovative visualization techniques to deepen insights and support global health initiatives.

=======
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(data.table)


#group by sum, line per region

# Set data directory
dir1 <- "../input/malaria-dataset/"

# Load datasets
estimated_numbers <- read.csv("data/archive 2/estimated_numbers.csv", stringsAsFactors = FALSE)
incidence <- read.csv("data/archive 2/incidence_per_1000_pop_at_risk.csv", stringsAsFactors = FALSE)
reported <- read.csv("data/archive 2/reported_numbers.csv", stringsAsFactors = FALSE)

# Merge reported and incidence datasets
data <- reported %>% 
  left_join(incidence, by = c("Country", "Year", "WHO.Region")) %>% 
  rename("No.of.cases" = "No..of.cases.x",
         "No.of.cases.per.1000.pop.at.risk" = "No..of.cases.y")

# Set plot size for better readability
options(repr.plot.width = 14, repr.plot.height = 10)

# Boxplot of global deaths by year (non-faceted)
reported %>% 
  drop_na(No..of.deaths) %>% 
  mutate(Year = as.factor(Year)) %>%
  ggplot(aes(x = Year, y = No..of.deaths)) +
  geom_boxplot(fill = "#56B4E9", alpha = 0.7) +
  labs(title = "Distribution of Reported Malaria Deaths by Year (Global)",
       y = "Number of Deaths", x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 🟠 Scatter plot of deaths over time, faceted by WHO Region
reported %>% 
  drop_na(No..of.deaths) %>% 
  mutate(Year = as.factor(Year)) %>%
  ggplot(aes(x = Year, y = No..of.deaths)) +
  geom_point(color = "#D55E00", alpha = 0.7) +
  labs(title = "Reported Malaria Deaths Over Time by WHO Region",
       y = "Number of Deaths", x = "Year") +
  facet_wrap(~ WHO.Region, scales = "free") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

>>>>>>> Stashed changes
```{r}
# Group by WHO Region and Year, then sum the deaths
deaths_by_region <- reported %>%
  drop_na(No..of.deaths) %>%
  group_by(WHO.Region, Year) %>%
  summarise(total_deaths = sum(No..of.deaths), .groups = 'drop')

# Line plot: deaths over time, one line per WHO region
ggplot(deaths_by_region, aes(x = Year, y = total_deaths, color = WHO.Region)) +
  geom_line(size = 1.2) +
  labs(title = "Total Reported Malaria Deaths Over Time by WHO Region",
       y = "Total Deaths", x = "Year", color = "WHO Region") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```
