---
title: "Malaria Cases: 2000-2017"
execute: 
  echo: true      # change to true to show the code
code-fold: false  # change to true to fold the code chunk
---

# Introduction

Malaria is a life-threatening disease caused by infected female *Anopheles* mosquitoes. This analysis explores malaria cases from 2000 to 2017 globally and by regions (Americas, Africa, Eastern Mediterranean, Europe, South-East Asia, Western Pacific), focusing on trends in case numbers and death rates.

# Research Question, Aims, and Methodology

## Research Question

**How has global malaria cases and reported death cases changed over time and across different regions from 2000 to 2017?**

## Aims

-   To visualize trends in malaria cases and deaths globally from 2000 onwards.
-   To analyze regional differences and identify patterns or anomalies over time.
-   To better understand where malaria remains a major public health challenge.

## Methodology

-   **Dataset:** WHO malaria data (reported cases and deaths by country and year).
-   **Visualization Methods:**
    -   Scatter plots to show temporal trends.
    -   Dynamic choropleth maps to demonstrate spatial changes.

# Dataset Overview

-   **Total Rows:** 1,944
-   **Total Columns:** 5
    -   `Country` (Categorical): 190 unique country names.
    -   `Year` (Numerical): Reporting year (2000–2017).
    -   `No. of cases` (Numerical): Reported malaria cases (0 to 15.1 million).
    -   `No. of deaths` (Numerical): Reported malaria deaths (0 to 39,054).
    -   `WHO Region` (Categorical): 6 regions (Africa, Americas, Eastern Mediterranean, Europe, South-East Asia, Western Pacific).

# Results

## Total Reported Malaria Cases Over Time

```{r malaria-cases-over-time}
reported$Year <- as.factor(reported$Year)
reported %>%
  group_by(Year) %>%
  summarize(total_cases = sum(No..of.cases, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = total_cases, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Total Reported Malaria Cases Over Time",
       x = "Year",
       y = "Total Cases") +
  theme_minimal()

```

## Distribution of Reported Malaria Deaths by Year (Global)

```{r malaria-deaths-global}

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(data.table)

# Set data directory
dir1 <- "../input/malaria-dataset/"

# Load datasets
estimated_numbers <- read.csv("data/archive 2/estimated_numbers.csv", stringsAsFactors = FALSE)
incidence <- read.csv("data/archive 2/incidence_per_1000_pop_at_risk.csv", stringsAsFactors = FALSE)
reported <- read.csv("data/archive 2/reported_numbers.csv", stringsAsFactors = FALSE)

# Merge reported and incidence datasets
data <- reported %>% 
  left_join(incidence, by = c("Country", "Year", "WHO.Region")) %>% 
  rename("No.of.cases" = "No..of.cases.x",
         "No.of.cases.per.1000.pop.at.risk" = "No..of.cases.y")

# Set plot size for better readability
options(repr.plot.width = 14, repr.plot.height = 10)

# Boxplot of global deaths by year (non-faceted)
reported %>% 
  drop_na(No..of.deaths) %>% 
  mutate(Year = as.factor(Year)) %>%
  ggplot(aes(x = Year, y = No..of.deaths)) +
  geom_boxplot(fill = "#56B4E9", alpha = 0.7) +
  labs(title = "Distribution of Reported Malaria Deaths by Year (Global)",
       y = "Number of Deaths", x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(data.table)

# Set data directory
dir1 <- "../input/malaria-dataset/"

# Load datasets
estimated_numbers <- read.csv("../../data/archive 2/estimated_numbers.csv", stringsAsFactors = FALSE)
incidence <- read.csv("../../data/archive 2/incidence_per_1000_pop_at_risk.csv", stringsAsFactors = FALSE)
reported <- read.csv("../../data/archive 2/reported_numbers.csv", stringsAsFactors = FALSE)

# Merge reported and incidence datasets
data <- reported %>% 
  left_join(incidence, by = c("Country", "Year", "WHO.Region")) %>% 
  rename("No.of.cases" = "No..of.cases.x",
         "No.of.cases.per.1000.pop.at.risk" = "No..of.cases.y")

# Set plot size for better readability
options(repr.plot.width = 14, repr.plot.height = 10)

# Boxplot of global deaths by year (non-faceted)
reported %>% 
  drop_na(No..of.deaths) %>% 
  mutate(Year = as.factor(Year)) %>%
  ggplot(aes(x = Year, y = No..of.deaths)) +
  geom_boxplot(fill = "#56B4E9", alpha = 0.7) +
  labs(title = "Distribution of Reported Malaria Deaths by Year (Global)",
       y = "Number of Deaths", x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

## Reported Malaria Deaths Over Time by Region

```{r malaria-deaths-region}
library(tidyverse)
library(ggplot2)
library(data.table)

# 🟠 Scatter plot of deaths over time, faceted by WHO Region
reported %>% 
  drop_na(No..of.deaths) %>% 
  mutate(Year = as.factor(Year)) %>%
  ggplot(aes(x = Year, y = No..of.deaths)) +
  geom_point(color = "#D55E00", alpha = 0.7) +
  labs(title = "Reported Malaria Deaths Over Time by WHO Region",
       y = "Number of Deaths", x = "Year") +
  facet_wrap(~ WHO.Region, scales = "free") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

## Total Reported Malaria Deaths Over Time by Region

```{r total-deaths-region}

# Set data directory
dir1 <- "../input/malaria-dataset/"

# Load datasets
estimated_numbers <- read.csv("data/archive 2/estimated_numbers.csv", stringsAsFactors = FALSE)
incidence <- read.csv("data/archive 2/incidence_per_1000_pop_at_risk.csv", stringsAsFactors = FALSE)
reported <- read.csv("data/archive 2/reported_numbers.csv", stringsAsFactors = FALSE)

# Merge reported and incidence datasets
data <- reported %>% 
  left_join(incidence, by = c("Country", "Year", "WHO.Region")) %>% 
  rename("No.of.cases" = "No..of.cases.x",
         "No.of.cases.per.1000.pop.at.risk" = "No..of.cases.y")

# Set plot size
options(repr.plot.width = 14, repr.plot.height = 10)

# 📈 Line graph of deaths over time by WHO Region
reported %>% 
  drop_na(No..of.deaths) %>% 
  group_by(WHO.Region, Year) %>%  # <-- group by region and year
  summarize(total_deaths = sum(No..of.deaths, na.rm = TRUE)) %>% 
  ggplot(aes(x = Year, y = total_deaths, color = WHO.Region, group = WHO.Region)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Reported Malaria Deaths Over Time by WHO Region",
       x = "Year",
       y = "Total Reported Deaths",
       color = "WHO Region") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

## Reported Cases by Country Over Time

```{r cases-by-country}
library(plotly)
fig <- plot_ly(
  data = malaria_clean,
  locations = ~country,
  locationmode = 'country names',
  z = ~cases,
  text = ~paste(country, "<br>Cases:", cases),
  type = 'choropleth',
  frame = ~year,
  colorscale = 'Reds',
  marker = list(line = list(color = "gray", width = 0.5))
)

fig <- fig %>%
  layout(
    title = "Reported Malaria Cases Over Time",
   geo = list(showframe = FALSE,
               showcoastlines = FALSE,
               showcountries = TRUE,
               countrycolor = "gray",
               projection = list(type = "equirectangular"),
               bgcolor = "white",
               landcolor = "white")
  )

fig
```

## Malaria Cases and Deaths by Country Over Time

```{r cases-deaths-country}
malaria_clean <- reported %>%
  select(Country, Year, `No..of.cases`,`No..of.deaths`) %>%
  rename(
    country = Country,
    year = Year,
    cases = `No..of.cases`,
    deaths = `No..of.deaths`
  )

fig <- plot_ly() %>%
  # Choropleth layer for total cases
  add_trace(
    data = malaria_clean,
    type = "choropleth",
    locations = ~country,
    locationmode = "country names",
    z = ~cases,
    frame = ~year,
    colorscale = "Reds",
    marker = list(line = list(color = "gray", width = 0.5)),
    text = ~paste(
      country,
      "<br>Cases: ", format(cases, big.mark = ","),
      "<br>Deaths: ", format(deaths, big.mark = ",")
    ),
    hoverinfo = "text"
  ) %>%
  
  # Scattergeo layer for deaths as circle markers
  add_trace(
    data = malaria_clean,
    type = "scattergeo",
    locations = ~country,
    locationmode = "country names",
    frame = ~year,
    marker = list(
      size = ~sqrt(deaths) / 100,  # Scale marker size
      color = "black",
      opacity = 0.5,
      line = list(width = 0)
    ),
    hoverinfo = "none",  # Already shown in choropleth hover
    showlegend = FALSE
  ) %>%
  
  layout(
    title = "Malaria Cases and Deaths Over Time",
    geo = list(
      showframe = FALSE,
      showcoastlines = FALSE,
      showcountries = TRUE,
               countrycolor = "gray",
               projection = list(type = "equirectangular"),
               bgcolor = "white",
               landcolor = "white"
    ),
    coloraxis = list(
      colorbar = list(title = "Cases")
    )
   
  )
fig
```

# Future Work and Questions

-   Explore causes behind regional differences (e.g., healthcare access, climate, interventions).
-   Update the dataset with more recent years.
-   Investigate specific country case studies for deeper insights.
-   Analyze the relationship between malaria interventions and case/death trends.

# Conclusion

Teamwork and visual analytics have enabled a clearer understanding of malaria's global impact. Future studies should continue to leverage updated data and innovative visualization techniques to deepen insights and support global health initiatives.
